Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-02-14 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: terse text about qnm translation

% Modified gravity related
\newcommand{\pd}{\partial}
\newcommand{\dd}{{\rm d}}
\newcommand{\dV}{{\rm d}^{4}x \, \sqrt{-g} \,}

@@

\hs{AG will do this.}
\hs{Here we present a short review of~\cite{Brito:2018rfr,Ghosh:2021mrv} and
explain how we modified it to include the {\sc Parspec} parametrisation.}

@@

\hs{HS will do this.}

\begin{table*}[th]
\begin{tabular}{c | c c c c c c c c}
\hline
\hline
Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
\hline
sGB      & 4 & 0.0107 & 0.0044 & \dots & \dots &  \cite{Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.7$~km~\cite{Perkins:2021mhb} & \dots \\
dCS      & 4 & 3.1964 & 6.3619 & \dots & \dots &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~\cite{Silva:2020acr}   & \dots \\
cEFTofGR & 4 & $-0.3680$ $(-0.5131)$  & 1.9242 (1.7000) & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
qEFTofGR & 6 & \dots & \dots & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
\hline
\hline
\end{tabular}
\caption{Summary of the quasinormal modes calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
references from which we used the results from and the current best constraint
(if applicable).
%
In the entries for cEFTofGR, the first entries in the $\delta
\omega^{(i)}_{220}$, $\delta \tau^{(i)}_{220}$ are obtained by considering
\emph{only parity-even corrections} and the second entries by considering
\emph{only parity-odd corrections}. In these cases, the dimensionless parameters
$\lambda_{\rm e,o}$ are degenerate with the length scale $\ell$.
%
\hs{Add a column with the constraint (if any) that we placed.}
\hs{TODO: add which runs are done and which need to be done.}
}
\label{tab:ref_theories_qnms}
\end{table*}

\hs{We {\it must very explicitly} state our working hypothesis here:
%
\begin{itemize}
    \item That we include only the nonrotating, non-GR correction to the QNMs.
    \item That due to absence of isospectrality, when translating to the theory specific result,
    we {\it chose} to use the lowest damping, gravitational QNM of the theory.
\end{itemize}
%
We can come back to these in the Sec.~\ref{sec:discussion} as things that we
might want to revisit in the future.
}

\hs{In Table~\ref{tab:ref_theories_qnms} we summarize the relevant parameters for
the {\sc Parspec} implementation of each theory and give credit to the papers
which calculated the QNMs.}

>>>>

% Modified gravity related
\newcommand{\pd}{\partial}
\newcommand{\dd}{{\rm d}}
\newcommand{\ii}{{\rm i}}
\newcommand{\dV}{{\rm d}^{4}x \, \sqrt{-g} \,}

@@

\hs{AG will do this.}
\hs{Here we present a short review of~\cite{Brito:2018rfr,Ghosh:2021mrv} and
explain how we modified it to include the {\sc Parspec} parametrisation.}
\input{pseob_review}

@@

Let us now establish the connection between theory-agnostic framework of the
$\pSEOB$ waveform model and the theory-specific QNM calculations.
%
We begin by expanding $\omega$ and $\tau$, given by Eqs.~\eqref{eq:kerr_expansion} as,
%
\begin{subequations}
\begin{align}
M \omega_{220} &= \gamma (\delta\omega^{(0)} \omega^{(0)} + \chi \delta\omega^{(1)} \omega^{(1)})
+ \sum_{j=0}^{N} \chi^{j} \omega^{j}\,,
\\
\tau_{220}/M   &= \gamma (\delta\tau^{(0)} \tau^{(0)} + \chi \delta\tau^{(1)} \tau^{(1)})
+ \sum_{j=0}^{N} \chi^{j} \tau^{j}\,,
\end{align}
\label{eq:qnm_nongr_iso}
\end{subequations}
%
where we are already focusing on the dominant $\ell = m = 2$, $n=0$ QNM and
pulled out of the sum all beyond-GR corrections, rescriting ourselves to the nonspinning ($j=0$) and linear-order in spin ($j=1$)
corrections to the Kerr QNMs, encoded in the coefficients $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$.

Just as in GR where the coefficients $\omega^{j}$, $\tau^{j}$ are determined by
comparison against the Kerr QNMs, a theory-specific calculation predicts the
numerical values of $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$
%
This means that using the $\pSEOB$ waveform model with theory-specific informed
QNMs we reduce the parameters space of beyond-GR parameters to one, i.e., to
only $\gamma$, which in turn contains information about the length scale $\ell$.

Now as we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating black holes
is modified gravity theories come in two families depending on their transformation upon a parity
transformation: axial and polar. Which is the one we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work on a theory-by-theory case and in each theory perform a
a translation between the metric perturbations $h_{\mu\nu}$ in the Regger-Wheeler-Zerilli gauge and
connect it with the transverse-traceless gauge used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion in Ref.~\cite{Hui:2021cpm} for a counterexample).
and thus how axial and polar gravitational QNMs appear in the GW signal has to be answered in theory-by-theory.
%
This is outside the scope of this paper and here we take the more pragmatic approach
of simply choosing the \emph{least damped gravitational mode between the two parities.}
%
The justification is this is the mode which, if excited during the merger, is
the most likely to appear in the signal.

In Table~\ref{tab:ref_theories_qnms} we summarize the relevant parameters for
the {\sc Parspec} implementation of each theory and give credit to the papers
which calculated the QNMs.

\begin{table*}[th]
\begin{tabular}{c | c c c c c c c c}
\hline
\hline
Theory & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & QNM & Constraint & This work \\
\hline
sGB      & 4 & 0.0107 & 0.0044 & \dots & \dots &  \cite{Pierini:2021jxd} & $\ell_{\rm GB} \leqslant 1.7$~km~\cite{Perkins:2021mhb} (GW) & \dots \\
dCS      & 4 & 3.1964 & 6.3619 & \dots & \dots &  \cite{Wagle:2021tam}   & $\ell_{\rm CS} \leqslant 8.5$~km~\cite{Silva:2020acr} (EM+GW)  & \dots \\
cEFTofGR & 4 & $-0.3680$ $(-0.5131)$  & 1.9242 (1.7000) & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
qEFTofGR & 6 & \dots & \dots & \dots & \dots & \cite{Cano:2021myl} & --  & \dots \\
\hline
\hline
\end{tabular}
\caption{Summary of the quasinormal modes calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(i)}_{220}$ and decay time $\delta \tau^{(i)}_{220}$, the
references from which we used the results from and the current best constraint
(if applicable).
%
In the entries for cEFTofGR, the first entries in the $\delta
\omega^{(i)}_{220}$, $\delta \tau^{(i)}_{220}$ are obtained by considering
\emph{only parity-even corrections} and the second entries by considering
\emph{only parity-odd corrections}. In these cases, the dimensionless parameters
$\lambda_{\rm e,o}$ are degenerate with the length scale $\ell$.
%
\hs{Add a column with the constraint (if any) that we placed.}
\hs{TODO: add which runs are done and which need to be done.}
}
\label{tab:ref_theories_qnms}
\end{table*}

% \hs{We {\it must very explicitly} state our working hypothesis here:
% %
% \begin{itemize}
%     \item That we include only the nonrotating, non-GR correction to the QNMs.
%     \item That due to absence of isospectrality, when translating to the theory specific result,
%     we {\it chose} to use the lowest damping, gravitational QNM of the theory.
% \end{itemize}
% %
% We can come back to these in the Sec.~\ref{sec:discussion} as things that we
% might want to revisit in the future.
% }

% \hs{In Table~\ref{tab:ref_theories_qnms} we summarize the relevant parameters for
% the {\sc Parspec} implementation of each theory and give credit to the papers
% which calculated the QNMs.}
